Natural variation in teosinte at the domestication locus \emph{teosinte branched1} (\emph{tb1})

### 2014-05-20 manuscript.tex
Edits to abstract

The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of lateral organ growth, is a major QTL involved in branching differences between maize and its wild progenitor, teosinte. Previous work has shown that the insertion of a transposable element (\emph{Hopscotch}) upstream of \emph{tb1} enhances its expression, causing much of the reduction in tillering observed in domesticated maize. Observations of the maize \emph{tb1} allele in teosinte individuals, coupled with estimates of the age of insertion of the \emph{Hopscotch} element, led us to investigate the prevalence and potential role of \emph{tb1} in teosinte. Results from genotyping across many natural populations suggest that the \emph{Hopscotch} element is segregating at a higher than expected frequency in a number of populations of two subspecies of teosinte, \emph{Zea mays} ssp. \emph{parviglumis} and \emph{Zea mays} ssp. \emph{mexicana}. Analysis of linkage disequilibrium between the \emph{Hopscotch} element and variation in surrounding regions does not support a hypothesis of recent introgression from maize into teosinte, and we find no evidence of environmental correlations that might suggest recent selection. Finally, two greenhouse experiments with \emph{Zea mays} ssp. \emph{parviglumis} do not suggest an important role for \emph{tb1} in controlling tillering in natural populations of this subspecies. Our findings suggest that the role of the \emph{Hopscotch} in tillering in teosinte is not as straightforward as is in domesticated maize, and that other loci may play a role in observed variation in this trait.

>>>>

The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of lateral organ growth, is a major QTL involved in branching differences between maize and its wild progenitor, teosinte. Previous work has shown that the insertion of a transposable element (\emph{Hopscotch}) upstream of \emph{tb1} enhances the gene's expression, causing much of the reduction in tillering observed in domesticated maize. Observations of the maize \emph{tb1} allele in teosinte individuals, coupled with estimates of the age of insertion of the \emph{Hopscotch} element, led us to investigate its prevalence and potential role in teosinte. Results from genotyping across many natural populations suggest that the \emph{Hopscotch} element is segregating at a higher than expected frequency in a number of populations of two subspecies of teosinte, \emph{Zea mays} ssp. \emph{parviglumis} and \emph{Zea mays} ssp. \emph{mexicana}. Analysis of linkage disequilibrium between the \emph{Hopscotch} element and variation in surrounding regions does not support a hypothesis of recent introgression from maize into teosinte.  Population genetic signatures, however, are consistent with selection on this locus and suggest the \emph{Hopscotch} insertion at \emph{tb1} may play an important ecological role in teosinte. Finally, two greenhouse experiments with \emph{Zea mays} ssp. \emph{parviglumis} do not suggest \emph{tb1} controls tillering in natural populations of this subspecies. Our findings suggest that the role of the \emph{Hopscotch} in tillering in teosinte differs from domesticated maize, and that other loci may play a role in observed variation in this trait.
